Finding shortest and closest vectors in a lattice of Voronoi's first kind

### 2014-04-21 paper.tex
minor

The closest lattice point problem and the shortest vector problem have interested mathematicians and computer scientists due to their relationship with integer programming~\cite{Lenstra_integerprogramming1983,Kannan1987_fast_general_np,Babai1986}, the factoring of polynomials~\cite{Lenstra1982}, and cryptanalysis~\cite{Joux_toolbox_cryptanal1998,NyguyenStern_two_faces_crypto,Micciancio_lattice_based_post_quantum_crypto}.  
Solutions of the closest lattice point problem also have engineering applications.  For example, if a lattice is used as a vector quantiser then the closest lattice point corresponds to the minimum distortion point~\cite{Conway1983VoronoiCodes,Conway1982VoronoiRegions,Conway1982FastQuantDec}.  If the lattice is used as a code, then the closest lattice point corresponds to what is called \emph{lattice decoding}\index{lattice decoding} and has been shown to yield arbitrarily good codes~\cite{Erex2004_lattice_decoding,Erez2005}.  The closest lattice point problem also occurs in communications systems involving multiple antennas~\cite{Ryan2008,Wubben_2011}.  The unwrapping of phase data for location estimation can also be posed as a closest lattice point problem and this has been applied to the global positioning system~\cite{Teunissen_GPS_1995,Hassibi_GPS_1998}.  The problem has also found applications to circular statistics~\cite{McKilliam_mean_dir_est_sq_arc_length2010}, single frequency estimation~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}, and related signal processing problems~\cite{McKilliam2007,Clarkson2007,McKilliam2009IndentifiabliltyAliasingPolyphase,Quinn_sparse_noisy_SSP_2012}.

@@

\begin{proposition}\label{eq:latticepointsinvorcvell}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice.  For $r\in \reals$ let $\ceil{r}$ denote the smallest integer strictly larger than $r$.  Let $t \in \reals^m$.  The number of lattice points inside the scaled and translated Voronoi cell $r \vor(\Lambda) + t$ is at most $\ceil{r}^n$.
\end{proposition}
\begin{proof}
It is convenient to modify the boundary of the Voronoi cell so that it tessellates $\reals^m$ under translations by $\Lambda$.  With this aim we let $V \subset \vor(\Lambda)$ contain all those points from the interior of $\vor(\Lambda)$, but have opposing faces open and closed.  That is, if $x \in V$ is on the boundary of $V$ then $-x \notin V$.  With this definition it can be asserted that $V$ tessellates $\reals^m$ under translations by $\Lambda$.  That is, $\reals^{m} = \cup_{x \in \Lambda}(V + x)$ and the intersection $(V + x)\cap(V+y)$ is empty for distinct lattice points $x$ and $y$~\cite[Sec.~2.2]{McKilliam2010thesis}.  Now, for positive integer $k$, the scaled and translated cell $kV + t$ contains precisely one coset representative for each element of the quotient group $\Lambda/k\Lambda$~\cite[Sec.~2.4]{McKilliam2010thesis}.  There are $k^n$ coset representatives.  Thus, the number of lattice points inside $r \vor(\Lambda) + t \subset \ceil{r}V + t$ is at most $\ceil{r}^n$.
\end{proof}

@@

Let $\Lambda$ be a lattice in $\reals^m$ and let $y \in \reals^m$. A simple method to compute the lattice point $x \in \Lambda$ closest to $y$ is as follows.  Let $x_0$ be some lattice point from $\Lambda$, for example the origin.  Consider the following iteration,
\begin{align}\
x_{k+1} &= x_k + v_k \nonumber \\
v_k &= \arg\min_{ v \in \relevant(\Lambda) \cup \{\zerobf\} } \|y - x_k - v \|, \label{eq:relvectsminimslicer}
\end{align} 
where $\relevant(\Lambda) \cup \{\zerobf\}$ is the set of relevant vectors of $\Lambda$ including the origin.  The minimum over $\relevant(\Lambda) \cup \{\zerobf\}$ may not be unique, that is, there may be multiple vectors from $\relevant(\Lambda) \cup \{\zerobf\}$ that are closest to $y - x_k$.  In this case, any one of the minimisers may be chosen.  The results that we will describe do not depend on this choice. We make the following straightforward propositions.

>>>>

The closest lattice point problem and the shortest vector problem have interested mathematicians and computer scientists due to their relationship with integer programming~\cite{Lenstra_integerprogramming1983,Kannan1987_fast_general_np,Babai1986}, the factoring of polynomials~\cite{Lenstra1982}, and cryptanalysis~\cite{Joux_toolbox_cryptanal1998,NyguyenStern_two_faces_crypto,Micciancio_lattice_based_post_quantum_crypto}.  
Solutions of the closest lattice point problem have engineering applications.  For example, if a lattice is used as a vector quantiser then the closest lattice point corresponds to the minimum distortion point~\cite{Conway1983VoronoiCodes,Conway1982VoronoiRegions,Conway1982FastQuantDec}.  If the lattice is used as a code, then the closest lattice point corresponds to what is called \emph{lattice decoding}\index{lattice decoding} and has been shown to yield arbitrarily good codes~\cite{Erex2004_lattice_decoding,Erez2005}.  The closest lattice point problem also occurs in communications systems involving multiple antennas~\cite{Ryan2008,Wubben_2011}.  The unwrapping of phase data for location estimation can also be posed as a closest lattice point problem and this has been applied to the global positioning system~\cite{Teunissen_GPS_1995,Hassibi_GPS_1998}.  The problem has also found applications to circular statistics~\cite{McKilliam_mean_dir_est_sq_arc_length2010}, single frequency estimation~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}, and related signal processing problems~\cite{McKilliam2007,Clarkson2007,McKilliam2009IndentifiabliltyAliasingPolyphase,Quinn_sparse_noisy_SSP_2012}.

@@

\begin{proposition}\label{eq:latticepointsinvorcvell}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice.  For $r\in \reals$ let $\ceil{r}$ denote the smallest integer strictly larger than $r$.  Let $t \in \reals^m$.  The number of lattice points inside the scaled and translated Voronoi cell $r \vor(\Lambda) + t$ is at most $\ceil{r}^n$.
\end{proposition}
\begin{proof}
It is convenient to modify the boundary of the Voronoi cell so that it tessellates $\reals^m$ under translations by $\Lambda$.  With this aim we let $V \subset \vor(\Lambda)$ contain all those points from the interior of $\vor(\Lambda)$, but have opposing faces open and closed.  That is, if $x \in V$ is on the boundary of $V$ then $-x \notin V$.  With this definition it can be asserted that $V$ tessellates $\reals^m$ under translations by $\Lambda$.  That is, $\reals^{m} = \cup_{x \in \Lambda}(V + x)$ and the intersection $(V + x)\cap(V+y)$ is empty for distinct lattice points $x$ and $y$.  Now, for positive integer $k$, the scaled and translated cell $kV + t$ contains precisely one coset representative for each element of the quotient group $\Lambda/k\Lambda$~\cite[Sec.~2.4]{McKilliam2010thesis}.  There are $k^n$ coset representatives.  Thus, the number of lattice points inside $r \vor(\Lambda) + t \subset \ceil{r}V + t$ is at most $\ceil{r}^n$.
\end{proof}

@@

Let $\Lambda$ be a lattice in $\reals^m$ and let $y \in \reals^m$. A simple method to compute a lattice point $x \in \Lambda$ closest to $y$ is as follows.  Let $x_0$ be some lattice point from $\Lambda$, for example the origin.  Consider the following iteration,
\begin{align}
x_{k+1} &= x_k + v_k \nonumber \\
v_k &= \arg\min_{ v \in \relevant(\Lambda) \cup \{\zerobf\} } \|y - x_k - v \|, \label{eq:relvectsminimslicer}
\end{align} 
where $\relevant(\Lambda) \cup \{\zerobf\}$ is the set of relevant vectors of $\Lambda$ including the origin.  The minimum over $\relevant(\Lambda) \cup \{\zerobf\}$ may not be unique, that is, there may be multiple vectors from $\relevant(\Lambda) \cup \{\zerobf\}$ that are closest to $y - x_k$.  In this case, any one of the minimisers may be chosen.  The results that we will describe do not depend on this choice. We make the following straightforward propositions.

### 2014-04-21 talk/presentation.tex
minor

\title{Finding shortest and closest vectors in a lattice of Voronoi's first kind}
\author{Robby McKilliam, Alex Grant, and I. Vaughan L. Clarkson \\ 
\vspace{0.5cm}
{\small
Institute for Telecommunications Research \\ University of South Australia 
\\ \vspace{0.2cm}
School of Information Technology and Electrical Engineering \\ The University of Queensland
}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics{figs/lattice-1.mps}
		\caption{A 2-dimensional lattice.}
		\label{fig:latfig1}
\end{figure}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics{figs/lattice-1.mps}
		\caption{A 2-dimensional lattice.}
		\label{fig:latfig2}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics{figs/latticewithspherepacking-1.mps}
		\caption{A 2-dimensional lattice.  There are 4 short vectors.}
		\label{fig:latfigandpacking1}
\end{figure}
}

\frame{
\frametitle{The shortest vector problem}
\begin{itemize}
\item Computing a short vector is called the \term{shortest vector problem}.
\item Applications in cryptography and number theory.
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item For example, short vectors are easy to find in the \term{root lattices} $\ints^n$, $A_n$ and $D_n$.
\item We will show that the problem is relatively easy to solve for lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

\frame{ 
\begin{figure}[tp]
	\centering 
		\includegraphics{figs/voronoicell-1.mps}
		\caption{A 2-dimensional lattice and its Voronoi cell.}
		\label{fig:latfigandpacking1} 
\end{figure}
}

@@

\frame{
\begin{figure}[tp] 
	\centering 
		\includegraphics{figs/relevantvectors-1.mps}  
		\caption{A 2-dimensional lattice with 6 relevant vectors.}
		\label{fig:latfigandpacking1}
\end{figure}
}

@@

\frame{
\begin{figure}[tp] 
	\centering 
		\includegraphics{figs/relevantvectorsandspherepacking-1.mps}   
		\caption{A lattice with 6 relevant vectors and 4 short vectors.}
		\label{fig:latfigandpacking1}
\end{figure}
}

\frame{
\frametitle{Closest lattice points}

@@

\begin{itemize}
\item This is called the \term{closest lattice point problem} and a solution is called a \term{closest lattice point} to $y$
\end{itemize}

@@

@@

\section{Examples}

@@

\begin{itemize}
\item Can we efficiently decide whether a lattice is of Voronoi's first kind?
\item Can we efficiently find an obtuse superbase if it exists?
\item Can a similar approach be taken to solve the \term{nearest lattice point problem}?
\end{itemize}

@@

\frame{
START TALKING ABOUT CLOSEST POINT ALGORITHM
}

@@

\begin{itemize}
\item Can good codes or quantisers be constructed from lattices of Voronoi's first kind?
\item Signal processing applications, phase unwrapping, circular statistics, etc.
\end{itemize}

>>>>

\title{Finding shortest and closest vectors in a lattice of Voronoi's first kind}
\author{Robby McKilliam \\
\vspace{0.2cm}
Alex Grant \\ Vaughan Clarkson \\
\vspace{0.7cm}
{\small
Institute for Telecommunications Research \\ University of South Australia 
\\ \vspace{0.2cm}
School of Information Technology and Electrical Engineering \\ The University of Queensland
}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics{figs/lattice-1.mps}
		\caption{A 2-dimensional lattice.}
		%\label{fig:latfig1}
\end{figure}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics{figs/lattice-1.mps}
		\caption{A 2-dimensional lattice.}
		%\label{fig:latfig2}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics{figs/latticewithspherepacking-1.mps}
		\caption{A 2-dimensional lattice.  There are 4 short vectors.}
		%\label{fig:latfigandpacking1}
\end{figure}
}

\frame{
\frametitle{The shortest vector problem}
\begin{itemize}
\item Computing a short vector is called the \term{shortest vector problem}.
\item Applications in cryptography and number theory.
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item For example, short vectors are easy to find in the \term{root lattices} $\ints^n$, $A_n$, and $D_n$.
\item We will show that the problem is relatively easy to solve for lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

\frame{ 
\begin{figure}[tp]
	\centering 
		\includegraphics{figs/voronoicell-1.mps}
		\caption{A 2-dimensional lattice and its Voronoi cell.}
		%\label{fig:latfigandpacking1} 
\end{figure}
}

@@

\frame{
\begin{figure}[tp] 
	\centering 
		\includegraphics{figs/relevantvectors-1.mps}
		\caption{A 2-dimensional lattice with 6 relevant vectors.}
		%\label{fig:latfigandpacking1}
\end{figure}
}

@@

\frame{
\begin{figure}[tp] 
	\centering 
		\includegraphics{figs/relevantvectorsandspherepacking-1.mps}   
		\caption{A lattice with 6 relevant vectors and 4 short vectors.}
		%\label{fig:latfigandpacking1}
\end{figure}
}

\frame{
\frametitle{The closest lattice point problem}

@@

\begin{itemize}
\item This is called the \term{closest lattice point problem} and a solution is called a \term{closest lattice point} to $y$.
\end{itemize}

@@

\frame{
\begin{figure}[tp]  
	\centering 
		\includegraphics{figs/closestpoint-1.mps}   
		\caption{The closest lattice point.}
		%\label{fig:latfigandpacking1} 
\end{figure}
}

\frame{
\frametitle{The closest lattice point problem}
\begin{itemize}
\item The lattice point $x \in \Lambda$ is closest to $y \in \reals^m$ if and only if
\[
y \in \vor(\Lambda) + x.
\]
\item Equivalently $x$ is closest to $y$ if and only if
\[
\dotprod{(y - x)}{v} \leq \tfrac{1}{2} \dotprod{v}{v}
\]
for all $v \in \relevant(\Lambda)$.
\end{itemize}
}

\frame{
\frametitle{The closest lattice point problem}
Applications to:
\begin{itemize}
\item coding and quantisation,
\item multi-antenna communications (MIMO),
\item unwrapping of phase data for electronic distance measurement in GPS and surveying,
\item single frequency estimation,
\item polynomial phase estimation,
\item circular statistics.
\end{itemize}
}

\frame{
\frametitle{The closest lattice point problem}
\begin{itemize}
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item For example, fast algorithms exist for the \term{root lattices} $\ints^n$, $A_n$, and $D_n$.
\item We will describe a fast algorithm to compute a closest point in lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

%\section{Examples}

@@

\begin{itemize}
\item Can we efficiently decide whether a lattice is of Voronoi's first kind?
\item Can we efficiently find an obtuse superbase if it exists?
\item Can a similar approach be taken to solve the \term{closest lattice point problem}?
\end{itemize}

@@

\section{A series of relevant vectors}

\frame{
\frametitle{A series of relevant vectors}

Let $x_0$ be some lattice point from $\Lambda$ and consider the following iteration,
\begin{align*}
x_{k+1} &= x_k + v_k \\
v_k &= \arg\min_{ v \in \relevant(\Lambda) \cup \{\zerobf\} } \|y - x_k - v \|
\end{align*} 
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/seriesofrelevantvectors-1.mps}
		\caption{Computing a closest point by a series of relevant vectors.}
		%\label{fig:graphcut4}
\end{figure}
}

\newtheorem{proposition}{Proposition}

\frame{
\frametitle{A series of relevant vectors}
 \begin{proposition}
 There is a finite number $K$ such that 
\[
x_K, x_{K+1}, x_{K+2}, \dots
\]
 are all closest points to $y$.
 \end{proposition}
}

\frame{
\frametitle{A series of relevant vectors}
\begin{itemize}
\item The number $K$ of iterations might be large.
\item The initial point $x_0$ affects $K$.
\item Minimising over the set of relevant vectors, that is computing
\[
\arg\min_{ v \in \relevant(\Lambda) \cup \{\zerobf\} } \|y - x_k - v \|
\]
might be expensive.
\item There are as many as $2^{n+1}-2$ relevant vectors.
\end{itemize}
}

\frame{
\frametitle{A series of relevant vectors}
For a lattice of Voronoi's first kind:
\begin{itemize}
\item The intial point $x_0$ can be chosen to ensure that at most $K = n$ iterations is required before a closest point in found.
\item Minimisation over the set of relevant vector can be performed by computing a minimimum cut in a flow network.
\item Using known algorithms a minimum cut can be found in $O(n^3)$ operations.
\item In total $O(n^4)$ operations are required to compute a closest lattice point.
\end{itemize}
}

\frame{
\frametitle{A series of relevant vectors}
{\color{gray}For a lattice of Voronoi's first kind:}
\begin{itemize}
\item The intial point $x_0$ can be chosen to ensure that at most $K = n$ iterations is required before a closest point in found.
\item {\color{gray}Minimisation over the set of relevant vector can be performed by computing a minimimum cut in a flow network.}
\item {\color{gray}Using known algorithms a minimum cut can be found in $O(n^3)$ operations.}
\item {\color{gray}In total $O(n^4)$ operations are required to compute a closest lattice point.}
\end{itemize}
}

@@

\begin{itemize}
\item Can good codes or quantisers be constructed from lattices of Voronoi's first kind?
\item Do applications, such as global position, phase unwrapping, circular statistics, involve lattices of Voronoi's first kind?
\end{itemize}
